---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r data loading, echo = TRUE}
activity <- read.csv("activity.csv")
activity[,1] <- as.numeric(activity[ ,1])
activity[,2] <- as.Date(activity[ ,2], format="%Y-%m-%d")
activity[,3] <- as.numeric(activity[ ,3])
```

## What is mean total number of steps taken per day?
1. Calculate the number of steps taken per day.
```{r steps per day calc, echo = TRUE}
## Subset data to exclude missing values
na_log <- is.na(activity[,1])
activity_sub <- activity[!na_log, ]

## Sum number of steps per day
day_sum <- tapply(activity_sub$steps, activity_sub$date, sum)
```

2. Make a histogram of total number of steps taken each day
``` {r steps per day histogram, echo = TRUE}
hist(day_sum, col="red", main="Histogram of Steps Taken Each Day", xlab="Steps")
```

3. Calculate and report the mean and median number of steps taken per day
``` {r steps mean and median calc, echo = TRUE}
day_sum_mean <- format(mean(day_sum), scientific = FALSE)
day_sum_median <- format(median(day_sum), scientific = FALSE)
```

The mean number of steps taken per day is `r I(day_sum_mean)` and the median number of steps taken per day is `r I(day_sum_median)`.

## What is the average daily activity pattern?
``` {r average daily activity pattern, echo = TRUE}
## Calculate the mean number of steps taken during each interval and plot
interval_mean <- aggregate(activity_sub$steps, list(interval=activity_sub$interval), mean)
plot(interval_mean$interval, interval_mean$x, type ="l", xlab="Interval", ylab="Steps", main="Average Number of Steps Taken for each 5 minute interval")

## Determine the interval with the highest mean number of steps taken
interval_max <- interval_mean$interval[interval_mean$x==max(interval_mean$x)]
```

The 5 minute interval that contains the highest number of steps on average across all days is `r I(interval_max)`.

## Imputing missing values
``` {r missing value calc, echo = TRUE}
na_row <- nrow(activity[is.na(activity$steps), ])
```

There are `r I(na_row)` number of missing values in the dataset.

Each NA value will be replaced with the average number of steps for that interval taken across all days.

``` {r NA fill-in, echo = TRUE}
## Create a new dataset with NA values filled in
activity_filled <- activity
for (i in 1:nrow(activity_filled)) {
     if (is.na(activity_filled[i,1])) {
          activity_filled[i,1] <- interval_mean$x[interval_mean$interval==activity_filled[i,3]]
     }
}

## Make a new histogram of the total number of steps taken each day
day_sum_filled <- tapply(activity_filled$steps, activity_filled$date, sum)
hist(day_sum_filled, col="red", main="Histogram of Steps Taken Each Day", xlab="Steps")

## Calculate the new mean and median
day_sum_filled_mean <- format(mean(day_sum_filled), scientific = FALSE)
day_sum_filled_median <- format(median(day_sum_filled), scientific = FALSE)
```

The new mean number of steps taken per day is `r I(day_sum_filled_mean)` and the new median number of steps taken per day is `r I(day_sum_filled_median)`.  When compared to the original mean of `r I(day_sum_mean)` and original median of `r I(day_sum_median)`, the mean did not change but the median did.  The difference between the two means and medians is fairly minimal.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

``` {r weekday factor, echo = TRUE}
activity_filled$weekday <- weekdays(activity_filled[,2])
for (i in 1:nrow(activity_filled)) {
     if (activity_filled$weekday[i] == "Saturday") {activity_filled$weekday[i] <- "weekend"}
     else if (activity_filled$weekday[i] == "Sunday") {activity_filled$weekday[i] <- "weekend"}
     else {activity_filled$weekday[i] <- "weekday"}
}
activity_filled$weekday <- as.factor(activity_filled$weekday)
```

2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

``` {r weekday panel plot, echo = TRUE}
## Compute the average number of steps taken per 5 minute interval for weekdays and weekends
### Subset all weekday rows and all weekend rows
activity_wkday <- activity_filled[activity_filled$weekday == "weekday", ]
activity_wkend <- activity_filled[activity_filled$weekday == "weekend", ]

### Calculate the mean of each interval for weekdays and weekends
interval_mean_wkday <- aggregate(list(steps=activity_wkday$steps), list(interval=activity_wkday$interval), mean)
interval_mean_wkend <- aggregate(list(steps=activity_wkend$steps), list(interval=activity_wkend$interval), mean)

### Combine weekday and weekend interval means into one data frame with a weekday factor variable
interval_mean_wkday$weekday <- as.factor("weekday")
interval_mean_wkend$weekday <- as.factor("weekend")
interval_mean_new <- rbind(interval_mean_wkday, interval_mean_wkend)

## Plot the data
library(lattice)
with(interval_mean_new, {xyplot(steps ~ interval | weekday, layout = c(1,2), type = "l")})
```
